Let's talk about Thurstone & Co. 

### 2024-10-23 index.qmd
section:issues; process:write

---
title: "Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation"
author:
  - name: 
      given: Jose Manuel
      family: Rivera Espejo
    orcid: 0000-0002-3088-2783
    url: https://www.uantwerpen.be/en/staff/jose-manuel-rivera-espejo_23166/
    email: JoseManuel.RiveraEspejo@uantwerpen.be
    corresponding: true
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Tine
      family: van Daal
      non-dropping-particle: van
    orcid: https://orcid.org/0000-0001-9398-9775
    url: https://www.uantwerpen.be/en/staff/tine-vandaal/
    email: tine.vandaal@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Sven
      family: De Maeyer
      non-dropping-particle: De
    orcid: 0000-0003-2888-1631
    url: https://www.uantwerpen.be/en/staff/sven-demaeyer/
    email: sven.demaeyer@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Steven
      family: Gillis
    orcid: 
    url: https://www.uantwerpen.be/nl/personeel/steven-gillis/
    email: steven.gillis@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Linguistics
        group: Centre for computational linguistics, psycholinguistics, and sociolinguistics (CLiPS)
funding: 
  statement: "The project was founded through the Research Fund of the University of Antwerp (BOF)."
keywords:
  - causal inference
  - probability
  - Thurstone
  - comparative judgement
  - directed acyclic graph
  - structural causal models
  - statistical modeling
abstract: |
  (to do)
key-points:
  - (to do)
date: last-modified
bibliography: references.bib
notebook-links: global
lightbox: true
citation: true
  # type: article-journal
  # container-title: "Psychometrika"
  # doi: ""
  # # url: https://example.com/summarizing-output
execute: 
  cache: true
  # eval: true
  echo: false
  # output: true
  # include: true
  warning: false
  error: false
  message: false
---

@@

<!-- 2. But what are the problems of DCJ as a method? -->
Nevertheless, despite the growing number of CJ studies, the unsystematic and fragmented research approaches employed in the literature have overlooked several critical issues concerning the method. These issues fall into three main categories: the disconnect between CJ's structural and measurement model, the over-reliance on Thurstone's Case 5 [-@Thurstone_1927] in its measurement model, and the role of the comparison algorithms and the number of comparisons on the method's reliability and validity. In the following sections, each issue will be discussed in detail, followed by the introduction of a theoretical model and its translation into a statistical model that addresses all three concerns simultaneously.

# Overlooked issues in CJ literature {#sec-theory-issues}

## The disconnect between structural and measurement model

<!-- 1. The BTL model is CJ's measurement model -->
In the CJ literature, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] serves as the measurement model for CJ because it specifies how latent variables are estimated from observed ones [@Hoyle_et_al_2023; @Kline_et_al_2023]. In a CJ study, multiple judges conduct several rounds of pairwise comparisons to evaluate the relative manifestation of a trait between a pair of stimuli. This evaluation results in a dichotomous outcome that indicates which stimulus is perceived to exhibit a higher degree of the trait. The BTL model then uses these observed outcomes to estimate scores that represent the latent trait of interest [@Pollitt_2012a; @Pollitt_2012b; @Whitehouse_2012; @Jones_2015; @vanDaal_et_al_2016; @Lesterhuis_2018; @Boonen_et_al_2020; @Bouwer_et_al_2023].

<!-- 2. But there is no structural model, only post estimation analyses -->
Moreover, in the CJ literature, it is common practice for BTL-generated scores, or their transformations, to undergo additional data analysis and hypothesis testing. Researchers have utilized these scores in independent analyses to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other scoring methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].

<!-- 3. But what is the problem with this? -->
<!-- However, the statistical literature cautions when using estimated scores for further independent data analysis and hypothesis testing. Since BTL-generated scores are parameter estimates that carry inherent uncertainty, separating the subsequent analyses from this uncertainty may lead to an artificial inflation of the precision and power of the analyses and tests. Consequently, this practice could increase the risk of committing a type I error, wherein a null hypothesis is incorrectly rejected [@McElreath_2020]. An appropriate approach to address this issue, consistent with the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023; @Kline_et_al_2023], is that these analyses should be conducted at the structural level rather than as independent analyses. Consequently, an approach that integrates the structural and measurement model should be followed. -->

<!-- However, there is often a lack of integration between the structural model, which captures the relationships among the variables, and the measurement model, which generates the CJ scores. This disconnect can lead to inaccurate inferences about the traits being measured. -->

## Case 5 and the measurement model

@@

## The efficiency and reliability of comparison algorithms

@@

## What is the solution

@@

## A theoretical model for the CJ {#sec-theory-scientific}

## From theory to statistical model {#sec-theory-statmodel}

>>>>

---
title: "Let's talk about Thurstone & Co. "
subtitle: "An information-theoretical model for comparative judgments, and its statistical translation"
author:
  - name: 
      given: Jose Manuel
      family: Rivera Espejo
    orcid: 0000-0002-3088-2783
    url: https://www.uantwerpen.be/en/staff/jose-manuel-rivera-espejo_23166/
    email: JoseManuel.RiveraEspejo@uantwerpen.be
    corresponding: true
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Tine
      family: van Daal
      non-dropping-particle: van
    orcid: https://orcid.org/0000-0001-9398-9775
    url: https://www.uantwerpen.be/en/staff/tine-vandaal/
    email: tine.vandaal@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Sven
      family: De Maeyer
      non-dropping-particle: De
    orcid: 0000-0003-2888-1631
    url: https://www.uantwerpen.be/en/staff/sven-demaeyer/
    email: sven.demaeyer@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Steven
      family: Gillis
    orcid: 
    url: https://www.uantwerpen.be/nl/personeel/steven-gillis/
    email: steven.gillis@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Linguistics
        group: Centre for computational linguistics, psycholinguistics, and sociolinguistics (CLiPS)
funding: 
  statement: "The project was founded through the Research Fund of the University of Antwerp (BOF)."
keywords:
  - causal inference
  - probability
  - Thurstone
  - comparative judgement
  - directed acyclic graph
  - structural causal models
  - statistical modeling
abstract: |
  (to do)
key-points:
  - (to do)
date: last-modified
bibliography: references.bib
notebook-links: global
lightbox: true
citation: true
  # type: article-journal
  # container-title: "Psychometrika"
  # doi: ""
  # # url: https://example.com/summarizing-output
execute: 
  cache: true
  # eval: true
  echo: false
  # output: true
  # include: true
  warning: false
  error: false
  message: false
---

@@

<!-- 2. But what are the problems of DCJ as a method? -->
Nevertheless, despite the growing number of studies on CJ, unsystematic and fragmented research approaches in the literature have left several critical issues unaddressed. These issues can be grouped into three main categories: the disconnect between CJ's structural and measurement model, the over-reliance on the assumptions of Thurstone's Case 5 [-@Thurstone_1927] in its measurement model, and the unclear role and impact of comparison algorithms and the number of comparisons on the method's reliability and validity. The following sections will discuss each of these issues in detail,, followed by the introduction of a theoretical model and its statistical translation, which aim to address all three concerns simultaneously.

# The three critical issues in CJ literature {#sec-theory-issues}

## The disconnect between structural and measurement models

<!-- 1. The BTL model is CJ's measurement model -->
In a typical CJ study, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] serves as the measurement model. This model specifies how latent variables are estimated from observed ones [@Hoyle_et_al_2023; @Kline_et_al_2023]. In the context of CJ, multiple judges engage in several rounds of pairwise comparisons to assess the relative manifestation of a trait between two stimuli. Each comparison generates a dichotomous outcome, indicating which stimulus is perceived to exhibit a higher degree of the trait. The BTL model then uses these observed outcomes to estimate scores that represent the latent trait of interest [@Pollitt_2012a; @Pollitt_2012b; @Whitehouse_2012; @Jones_2015; @vanDaal_et_al_2016; @Lesterhuis_2018; @Boonen_et_al_2020; @Bouwer_et_al_2023].

<!-- 2. But there is no structural model, only post estimation analyses -->
Moreover, researchers often use these BTL-generated scores or their transformations separately, applying additional analyses or hypothesis testing. The CJ literature shows that these scores have been employed to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other scoring methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].

<!-- 3. But what is the problem with this? and what is the solution? -->
However, the statistical literature warns against using estimated scores for separate analyses and hypothesis testing. BTL-generated scores are parameter estimates that inherently carry uncertainty. Neglecting this uncertainty can inflate the precision and statistical power in independent analyses and tests. This oversight, in turn, may increase the risk of committing a type I error, where the null hypothesis is incorrectly rejected [@McElreath_2020]. To mitigate this risk, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023; @Kline_et_al_2023] recommend conducting such analyses within the structural model of CJ rather than separately. Therefore, an integrated approach combining both structural and measurement models is advisable.

## The assumptions of Case 5 and the measurement model

@@

## The role and impact of comparison algorithms

@@

@@

## A theoretical model for CJ {#sec-theory-theoretical}

## From theory to statistics {#sec-theory-statistics}
